---
title: "DiD-Based Quantile Treatment Effects"
author: "Brantly Callaway"
date: today
format: html
bibliography: refs.bib
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig.width: 6
    fig.height: 4
vignette: >
  %\VignetteIndexEntry{DiD-Based Quantile Treatment Effects}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

The `qte` package provides six DiD-based estimators for the QTT. All share
the same `yname`/`gname`/`tname`/`idname` interface and support staggered
treatment adoption natively. They differ in their identification assumptions
and the restrictions those assumptions impose on the data-generating process.

This vignette describes each estimator conceptually and shows the minimal
call. For a full applied workflow with aggregations and event-study plots, see
`vignette("staggered-adoption")`.

```{r}
#| label: setup
#| message: false
library(qte)
set.seed(42)
data(mpdta, package = "did")
```

We use `mpdta` throughout: a balanced panel of 500 US counties observed
annually from 2003–2007, with `lemp` (log employment) as the outcome and
`first.treat` as the cohort variable (year of first minimum wage increase; 0
= never treated).

The common call structure for all six estimators is:

```r
estimator(
  yname  = "lemp",        # outcome
  gname  = "first.treat", # first treatment period (0 = never treated)
  tname  = "year",        # calendar time
  idname = "countyreal",  # unit id
  data   = mpdta,
  gt_type = "att",        # "att" or "qtt"
  biters  = 100
)
```

---

## Change in Changes — `cic()`

**Reference:** @athey-imbens-2006

**Identification assumption:** The rank of each unit in the untreated outcome
distribution is stable over time (rank invariance), and the support of the
treated group's pre-period distribution is contained in the support of the
untreated group's pre-period distribution (support condition). No functional
form for the outcome equation is assumed.

**When to use:** A flexible nonparametric alternative to linear DiD when the
parallel trends assumption is implausible in levels but rank stability is
credible. Works with panel or repeated cross sections.

**Key restriction:** Stronger than standard parallel trends — requires the
entire rank ordering to be preserved, not just the mean.

```{r}
#| label: cic-att
#| cache: true
res_cic <- cic(
  yname  = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)
summary(res_cic)
```

---

## Quantile Difference-in-Differences — `qdid()`

**Reference:** @callaway-li-oka-2018

**Identification assumption:** Distributional parallel trends — the entire
untreated potential outcome distribution shifts in parallel across time for
treated and untreated groups (not just the mean). Formally, $F_{Y(0)|D=1,t}$
shifts over time in the same way as $F_{Y(0)|D=0,t}$.

**When to use:** A direct distributional analog of standard DiD. Weaker than
CiC in that it does not require rank invariance, but stronger than mean DiD.
Works with panel or repeated cross sections.

**Key restriction:** Requires the full distributional shift to be parallel,
not just the mean shift; this is testable in pre-treatment periods.

```{r}
#| label: qdid-att
#| cache: true
res_qdid <- qdid(
  yname  = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)
summary(res_qdid)
```

---

## Distributional Difference-in-Differences — `ddid()`

**Reference:** @callaway-li-oka-2018

**Identification assumption:** Conditional distributional parallel trends —
the untreated potential outcome distribution shifts in parallel conditional on
pre-treatment covariates. Extends QDiD to allow covariate adjustment.

**When to use:** When distributional parallel trends holds only after
conditioning on covariates. The covariate adjustment is done via outcome
regression on the CDFs, making it more flexible than linear regression
adjustment for standard DiD.

**Key restriction:** Same spirit as QDiD but allows covariates. Requires
panel data.

```{r}
#| label: ddid-att
#| cache: true
res_ddid <- ddid(
  yname  = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)
summary(res_ddid)
```

---

## Mean Difference-in-Differences — `mdid()`

**Reference:** @callaway-li-2019

**Identification assumption:** Standard (mean) parallel trends — the mean of
the untreated potential outcome evolves in parallel for treated and untreated
groups. This is the same assumption as in standard DiD.

**When to use:** When only mean parallel trends is credible, not distributional
parallel trends. `mdid()` recovers the ATT under this weaker assumption and
can also return the QTT by combining mean DiD with a rank-based
counterfactual distribution.

**Key restriction:** Weakest distributional assumption of the four DiD
methods. The QTT estimate relies on additional rank-stability within the mean
DiD framework.

```{r}
#| label: mdid-att
#| cache: true
res_mdid <- mdid(
  yname  = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)
summary(res_mdid)
```

---

## Panel QTT — `panel_qtt()`

**Reference:** @callaway-li-2019

**Identification assumption:** Copula stability — the dependence structure
(copula) between the pre-treatment and post-treatment untreated potential
outcomes is the same for treated and untreated groups. This is a panel-data
assumption that uses three periods: two pre-treatment periods to estimate the
copula, and one post-treatment period.

**When to use:** When panel data with at least two pre-treatment periods per
cohort are available and the copula stability assumption is credible. Uniquely
identifies the individual-level rank mapping from pre to post, enabling
inference on the QTT under weaker distributional assumptions than CiC.

**Key restriction:** Requires panel data with sufficient pre-treatment
periods. The `pre_copula` argument controls whether a fixed (`"short"`) or
rolling (`"long"`, default) base period is used.

```{r}
#| label: panel-qtt
#| cache: true
res_pqtt <- panel_qtt(
  yname  = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "qtt", probs = seq(0.1, 0.9, 0.1), biters = 50
)
summary(res_pqtt)
```

---

## Lagged-Outcome Unconfoundedness — `lou_qte()`

**Identification assumption:** Unconfoundedness conditional on the lagged
outcome — treatment is as good as randomly assigned after conditioning on the
pre-treatment value of the outcome. This is a selection-on-observables
assumption, not a parallel trends assumption.

**When to use:** When the pre-treatment outcome is a strong proxy for the
unobserved confounders driving selection into treatment, and the parallel
trends assumption is implausible. Common in labor and health economics where
lagged earnings or health status are strong predictors of treatment.

**Key restriction:** Does not require parallel trends, but assumes that
conditioning on the lagged outcome absorbs all confounding. Requires panel
data.

```{r}
#| label: lou-att
#| cache: true
res_lou <- lou_qte(
  yname  = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)
summary(res_lou)
```

---

## Summary

| Function | Assumption | Panel required | Notes |
|---|---|---|---|
| `cic()` | Rank invariance + support | Optional | Nonparametric; strongest assumption |
| `qdid()` | Distributional parallel trends | Optional | DiD analog for distributions |
| `ddid()` | Conditional distributional PT | Yes | Adds covariate adjustment to QDiD |
| `mdid()` | Mean parallel trends | Optional | Weakest DiD assumption |
| `panel_qtt()` | Copula stability | Yes | Requires ≥ 2 pre-periods per cohort |
| `lou_qte()` | Lagged-outcome unconfoundedness | Yes | Not a DiD assumption |

All estimators return either a `pte_results` object (`gt_type = "att"`) or a
`pte_qtt` object (`gt_type = "qtt"`), and support `autoplot()` for plotting
overall curves and event-study summaries. See `vignette("staggered-adoption")`
for the full applied workflow.

## References
