---
title: "Quantile Treatment Effects in R"
author: "Brantly Callaway"
date: today
format: html
bibliography: refs.bib
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig.width: 6
    fig.height: 4
vignette: >
  %\VignetteIndexEntry{Quantile Treatment Effects in R}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## What are Quantile Treatment Effects?

The Average Treatment Effect (ATE) or Average Treatment Effect on the Treated
(ATT) summarizes the causal effect of a treatment by a single number: the mean
difference in outcomes between treated and untreated units. This is often a
natural quantity, but it can miss important heterogeneity. If a job training
program substantially raises earnings at the bottom of the distribution while
having little effect at the top, the ATE obscures this. If a minimum wage
increase compresses the wage distribution, an average effect conceals the
compression.

The Quantile Treatment Effect at level $\tau$ is $QTE(\tau) = Q_{Y(1)}(\tau) -
Q_{Y(0)}(\tau)$, where $Q_{Y(d)}(\tau)$ is the $\tau$-th quantile of the
potential outcome distribution under treatment status $d$. Mapping $QTE(\tau)$
across $\tau \in (0,1)$ traces how the treatment shifts the entire outcome
distribution. The Quantile Treatment Effect on the Treated ($QTT$) conditions
on the treated group.

This vignette demonstrates `unc_qte()`, the `qte` package's estimator for
cross-sectional settings (no panel data required). For DiD-based estimators
and staggered treatment adoption, see `vignette("panel-estimators")` and
`vignette("staggered-adoption")`.

```{r}
#| label: setup
#| message: false
library(qte)
library(ggplot2)
set.seed(42)
data(lalonde)
```

## Random assignment

Under random assignment, no covariate adjustment is needed: the unconditional
quantiles of the treated and control outcome distributions identify the QTE
directly. We use the experimental Lalonde data (`lalonde.exp`), where
treatment was randomly assigned.

```{r}
#| label: random-qte
#| cache: true
res_exp <- unc_qte(
  yname  = "re78",
  dname  = "treat",
  data   = lalonde.exp,
  target = "qte",
  probs  = seq(0.1, 0.9, 0.1),
  biters = 50
)
summary(res_exp)
```

```{r}
#| label: random-qte-plot
#| fig-alt: "QTE curve under random assignment"
autoplot(res_exp, ylab = "QTE (earnings, 1978)")
```

The confidence band indicates that the QTE is positive across much of the
distribution, with the effect somewhat larger in the lower quantiles.

## Selection on observables

When treatment is not randomly assigned, the unconfoundedness assumption
(conditional independence of potential outcomes given covariates) provides
identification. We use the observational Lalonde data (`lalonde.psid`), which
combines the treated group from the experiment with a comparison group from
the PSID.

`unc_qte()` supports three estimation methods:

- `"ipw"` — inverse probability weighting (propensity score re-weighting)
- `"or"` — outcome regression (conditional quantile regression)
- `"aipw"` — augmented IPW (doubly robust; recommended)

```{r}
#| label: obs-qte
#| cache: true
xf <- ~ age + I(age^2) + education + black + hispanic + married + nodegree

res_psid <- unc_qte(
  yname      = "re78",
  dname      = "treat",
  data       = lalonde.psid,
  xformla    = xf,
  est_method = "aipw",
  target     = "qte",
  probs      = seq(0.1, 0.9, 0.1),
  biters     = 100
)
summary(res_psid)
```

```{r}
#| label: obs-qte-plot
#| fig-alt: "QTE curve under selection on observables"
autoplot(res_psid, ylab = "QTE (earnings, 1978)")
```

## QTT under selection on observables

Setting `target = "qtt"` estimates the Quantile Treatment Effect on the
Treated — the distributional effect for the subpopulation that actually
received treatment.

```{r}
#| label: qtt
#| cache: true
res_qtt <- unc_qte(
  yname      = "re78",
  dname      = "treat",
  data       = lalonde.psid,
  xformla    = xf,
  est_method = "aipw",
  target     = "qtt",
  probs      = seq(0.1, 0.9, 0.1),
  biters     = 100
)
summary(res_qtt)
```

```{r}
#| label: qtt-plot
#| fig-alt: "QTT curve under selection on observables"
autoplot(res_qtt, ylab = "QTT (earnings, 1978)")
```

## Reading the output

`summary()` prints the overall ATT, the QTE/QTT at each quantile, and
standard errors with confidence intervals. `autoplot()` returns a `ggplot`
object so you can add layers:

```{r}
#| label: custom-plot
#| eval: false
autoplot(res_qtt) +
  ggplot2::labs(title = "QTT — Lalonde (observational)",
                subtitle = "AIPW with pre-treatment covariates")
```

By default `autoplot()` shows a uniform confidence band (`cband = TRUE`),
which provides simultaneous coverage over all quantile levels — a stronger
guarantee than pointwise intervals. Pass `cband = FALSE` for pointwise
intervals instead.

Standard errors are computed via the empirical bootstrap. The `biters`
argument controls the number of iterations (default 100). Parallel computation
is available via the `cl` argument.

## References
