---
title: "Change in Changes with Staggered Treatment Adoption"
author: "Brantly Callaway"
date: today
format: html
bibliography: refs.bib
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig.width: 6
    fig.height: 4
vignette: >
  %\VignetteIndexEntry{Change in Changes with Staggered Treatment Adoption}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The Change in Changes (CiC) estimator of @athey-imbens-2006 identifies the
Quantile Treatment Effect on the Treated (QTET) under a rank-invariance and
support condition, without requiring a linear functional form.  The original
paper considers a two-period, two-group setting.

In many empirical applications, treatment is adopted at different times by
different units — a setting called *staggered treatment adoption*.
The `cic()` function in the `qte` package extends CiC to this setting by
estimating a group-time ATT, $ATT(g, t)$, for every combination of treatment
cohort $g$ (the period a unit is first treated) and calendar time $t$, then
aggregating these into interpretable summary parameters following the approach
of @callaway-santanna-2021.

This vignette demonstrates `cic()` using the `mpdta` dataset from the `did`
package [@callaway-santanna-2021], which contains county-level employment and
population data used to study the effect of minimum wage increases.

```{r}
#| label: load-packages
#| message: false
#| warning: false
library(qte)
library(did)
library(ggplot2)
data(mpdta)
```

## Data

`mpdta` is a balanced panel of 500 US counties observed annually from 2003 to
2007.  The outcome is `lemp` (log county-level employment).  Counties are
grouped by `first.treat`, the year they first adopted a minimum wage increase
(2004, 2006, or 2007); counties with `first.treat == 0` never adopt.

```{r}
#| label: data-summary
table(mpdta$first.treat)
```

## Estimating ATT(g,t) and Aggregations

`cic()` with the default `gt_type = "att"` estimates the Average Treatment
Effect on the Treated for each group-time pair and aggregates them into an
overall ATT and an event-study (dynamic) summary.

```{r}
#| label: att-estimate
#| cache: true
set.seed(42)
res_att <- cic(
  yname   = "lemp",
  gname   = "first.treat",
  tname   = "year",
  idname  = "countyreal",
  data    = mpdta,
  biters  = 200
)
```

`summary()` prints the overall ATT and the event-study coefficients:

```{r}
#| label: att-summary
summary(res_att)
```

The overall ATT is small and not statistically significant at conventional
levels.  The event-study rows show the estimated effects at each relative
time period $e = t - g$.  Negative event times ($e < 0$) are pre-treatment
periods; we expect these to be near zero if the identifying assumption holds.
Positive event times show the dynamic treatment effect after adoption.

## Estimating the QTT Curve

With `gt_type = "qtt"`, `cic()` returns the full Quantile Treatment Effect
on the Treated curve at each aggregation level.  CDFs are mixed across
group-time cells first (using overall-ATT weights), then inverted at the
requested `probs` grid — this avoids the bias that would arise from averaging
scalar quantiles.

```{r}
#| label: qtt-estimate
#| cache: true
set.seed(42)
res_qtt <- cic(
  yname   = "lemp",
  gname   = "first.treat",
  tname   = "year",
  idname  = "countyreal",
  data    = mpdta,
  biters  = 200,
  gt_type = "qtt",
  probs   = seq(0.1, 0.9, by = 0.1)
)
```

`print()` shows the overall QTT curve with pointwise 95% confidence intervals:

```{r}
#| label: qtt-print
print(res_qtt)
```

The overall QTT is modest and not significantly different from zero across
the distribution, consistent with the near-zero overall ATT.

### Plotting the overall QTT curve

The `overall` element of a `pte_qtt` object is a plain data frame, making
it straightforward to plot with `ggplot2`:

```{r}
#| label: qtt-plot
ggplot(res_qtt$overall, aes(x = probs, y = qtt)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2),
                     labels = scales::percent_format(accuracy = 1)) +
  labs(
    x     = "Quantile",
    y     = "QTT (log employment)",
    title = "Overall QTT Curve — CiC with Staggered Adoption"
  ) +
  theme_bw()
```

### Dynamic QTT

The `dynamic` element contains the QTT curve at each event time.  Here we
plot the median QTT ($\tau = 0.5$) across event times as an event study:

```{r}
#| label: qtt-dynamic-plot
dyn_median <- subset(res_qtt$dynamic, probs == 0.5)

ggplot(dyn_median, aes(x = e, y = qtt)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = -0.5, linetype = "dotted") +
  labs(
    x     = "Event time (years relative to treatment)",
    y     = "Median QTT (log employment)",
    title = "Event Study — Median QTT"
  ) +
  theme_bw()
```

## Repeated Cross Sections

If the data are repeated cross sections rather than a panel, set
`panel = FALSE` and omit `idname`.  `cic()` will use the repeated
cross-sections CiC variant internally:

```{r}
#| label: rcs-example
#| eval: false
res_rcs <- cic(
  yname   = "lemp",
  gname   = "first.treat",
  tname   = "year",
  data    = mpdta,   # no idname
  panel   = FALSE,
  biters  = 200
)
```

## Note on Inference

Standard errors are computed via the empirical bootstrap, resampling
counties (panel) or observations (repeated cross sections) with replacement.
The default `biters = 100` is suitable for exploration; use `biters = 499`
or more for publication results.  Parallel computation is available via the
`cl` argument.

## References
