---
title: "Staggered Treatment Adoption with the qte Package"
author: "Brantly Callaway"
date: today
format: html
bibliography: refs.bib
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig.width: 6
    fig.height: 4
vignette: >
  %\VignetteIndexEntry{Staggered Treatment Adoption with the qte Package}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Introduction

In many empirical settings, units adopt treatment at different points in time
— a structure called *staggered treatment adoption*. For example, states may
adopt a policy in different years, or firms may receive an intervention at
different times during a study period.

All DiD-based estimators in `qte` support staggered adoption via the same
interface: specify the outcome (`yname`), the first-treatment-period cohort
variable (`gname`, with 0 for never-treated units), calendar time (`tname`),
and a unit identifier (`idname`). The package follows the approach of
@callaway-santanna-2021, computing group-time effects $ATT(g, t)$ for every
cohort $g$ and period $t$, then aggregating these into overall, group-specific,
and event-study (dynamic) summaries.

This vignette demonstrates the full applied workflow using the `mpdta` dataset.
For conceptual background on the identification assumptions behind each
estimator, see `vignette("did-estimators")`.

```{r}
#| label: setup
#| message: false
library(qte)
library(ggplot2)
set.seed(42)
data(mpdta, package = "did")
```

`mpdta` is a balanced panel of 500 US counties observed annually from 2003 to
2007. The outcome is `lemp` (log county-level employment). Counties are grouped
by `first.treat`, the year they first adopted a minimum wage increase (2004,
2006, or 2007); counties with `first.treat == 0` never adopt.

```{r}
#| label: data-summary
table(mpdta$first.treat)
```

---

## ATT estimation and aggregation

We begin with the Change in Changes estimator (`cic()`) using `gt_type = "att"`.

```{r}
#| label: att-estimate
res_att <- cic(
  yname  = "lemp",
  gname  = "first.treat",
  tname  = "year",
  idname = "countyreal",
  data   = mpdta,
  biters = 50
)
```

### Overall ATT and event-study summary

`summary()` prints the overall ATT and the event-study (dynamic) table.
Event times $e < 0$ are pre-treatment periods; $e \geq 0$ are post-treatment.

```{r}
#| label: att-summary
summary(res_att)
```

### Event-study plot

`autoplot()` dispatches on the `pte_results` class and produces an event-study
plot by default. Use `type = "dynamic"` explicitly:

```{r}
#| label: att-event-study
#| fig-alt: "Event-study plot of ATT estimates by event time"
autoplot(res_att, type = "dynamic")
```

Pre-treatment estimates near zero support the identifying assumption. The
post-treatment pattern shows the estimated ATT in each period after adoption.

---

## QTT curve estimation

Setting `gt_type = "qtt"` returns the full Quantile Treatment Effect on the
Treated curve at each aggregation level. CDFs are mixed across group-time
cells first (using overall-ATT weights), then inverted at the requested
`probs` grid — this avoids the bias that would arise from averaging scalar
quantiles across cells.

```{r}
#| label: qtt-estimate
res_qtt <- cic(
  yname   = "lemp",
  gname   = "first.treat",
  tname   = "year",
  idname  = "countyreal",
  data    = mpdta,
  biters  = 200,
  gt_type = "qtt",
  probs   = seq(0.1, 0.9, by = 0.1)
)
```

### Overall QTT curve

```{r}
#| label: qtt-plot
#| fig-alt: "Overall QTT curve from CiC"
autoplot(res_qtt)
```

### Dynamic QTT — event-study by quantile

`autoplot()` with `type = "dynamic"` plots the QTT at selected quantile(s)
across event times. The `plot_probs` argument selects which quantiles to show.

```{r}
#| label: qtt-dynamic-single
#| fig-alt: "Event-study plot for the median QTT"
autoplot(res_qtt, type = "dynamic", plot_probs = 0.5)
```

Multiple quantiles can be overlaid by passing a vector to `plot_probs` (values must be in the estimated `probs` grid):

```{r}
#| label: qtt-dynamic-multi
#| fig-alt: "Event-study plot for multiple quantiles of the QTT"
autoplot(res_qtt, type = "dynamic", plot_probs = c(0.1, 0.5, 0.9))
```

---

## ATT vs. QTT

Running the same estimator with both `gt_type = "att"` and `gt_type = "qtt"`
allows a direct comparison: is the distributional effect constant (QTT ≈ ATT
at all $\tau$), or does treatment shift the distribution in a non-uniform way?

The overall ATT from `res_att` is approximately equal to the overall mean of
the QTT curve from `res_qtt` when effects are roughly uniform. Deviations
indicate distributional heterogeneity.

---

## Comparing estimators

All six estimators share the same interface. Here we run three of them with
`gt_type = "att"` for a quick comparison of overall ATT estimates:

```{r}
#| label: compare-estimators

res_qdid <- qdid(
  yname = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)

res_mdid <- mdid(
  yname = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)

res_lou <- lou_qte(
  yname = "lemp", gname = "first.treat", tname = "year",
  idname = "countyreal", data = mpdta,
  gt_type = "att", biters = 50
)
```

```{r}
#| label: compare-table
overall_att <- function(res, label) {
  data.frame(
    estimator = label,
    att       = round(res$overall_results$att, 4),
    se        = round(res$overall_results$se,  4)
  )
}

do.call(rbind, list(
  overall_att(res_att,  "CiC"),
  overall_att(res_qdid, "QDiD"),
  overall_att(res_mdid, "MDiD"),
  overall_att(res_lou,  "Lagged-outcome unconfoundedness")
))
```

Estimates will generally differ because each estimator relies on a different
identifying assumption. Similarity across methods strengthens confidence in
the findings; divergence warrants investigation of which assumption is more
plausible for the application.

---

## Panel QTT and the `pre_copula` option

`panel_qtt()` requires at least two pre-treatment periods per cohort to
estimate the copula that is transferred forward. The `pre_copula` argument
controls how the base period is chosen:

- `"long"` (default): the base period expands with the event horizon, making
  the copula window match the length of the post-treatment period. Requires
  more pre-treatment periods but is more internally consistent at longer
  horizons.
- `"short"`: always uses the two periods immediately before treatment,
  regardless of event horizon.

```{r}
#| label: panel-qtt
res_pqtt <- panel_qtt(
  yname      = "lemp",
  gname      = "first.treat",
  tname      = "year",
  idname     = "countyreal",
  data       = mpdta,
  pre_copula = "long",
  gt_type    = "qtt",
  probs      = seq(0.1, 0.9, 0.1),
  biters     = 200
)
autoplot(res_pqtt)
```

---

## Repeated cross sections

If the data are repeated cross sections rather than a panel, set `panel = FALSE`
and omit `idname`. The CiC, QDiD, and MDiD estimators support repeated cross
sections; `ddid()`, `panel_qtt()`, and `lou_qte()` require panel data.

```{r}
#| label: rcs-example
#| eval: false
res_rcs <- cic(
  yname  = "lemp",
  gname  = "first.treat",
  tname  = "year",
  data   = mpdta,   # no idname
  panel  = FALSE,
  biters = 50
)
```

---

## Inference notes

Standard errors are computed via the empirical bootstrap, resampling units
(panel) or observations (repeated cross sections) with replacement. The
`biters` argument controls the number of iterations (default 100; use 499 or
more for publication results). Parallel computation is available via the `cl`
argument (`cl = 4` uses 4 cores).

The `cband = TRUE` default in `summary()` and `autoplot()` produces uniform
confidence bands (simultaneous coverage over all quantiles or event times).
Set `cband = FALSE` for pointwise intervals.

## References
